Expressive Array Constructs in an Embedded GPU Kernel Programming Language

### 2013-10-01 src2/arbb/haskell_arbb.tex
apa

% regiment,
As a research area, EDSLs and two-stage DSLs have been actively pursued for at least a
decade \cite{vertigo, wavescript} but are gaining steam
recently 
%\cite{sejits, stanford-ppl} 
\cite{copperhead, stanford-ppl} 
and are beginning to appear in
commercial products \cite{ArBB}.
%
Further, EDSL techniques have spread beyond their origin in 
 the programming languages community.
% 
For example, both Stanford's Parallel Programming Laboratory (PPL) and
the Berkeley Parlab are creating EDSLs as their flagship parallel
programming solutions for domains such as machine learning and
rendering \cite{stanford-ppl}.
  Moreover, EDSLs need not be hosted by esoteric research languages---Intel's
 ArBB embeds an array language in C++ and Berkeley's
Copperhead \cite{copperhead} generates CUDA from simple Python code.

>>>>

% regiment,
As a research area, EDSLs and two-stage DSLs have been actively pursued for at least a
decade \cite{VERTIGO, wavescript} but are gaining steam
recently 
%\cite{sejits, stanford-ppl} 
\cite{copperhead, stanford-ppl} 
and are beginning to appear in
commercial products \cite{ArBB}.
%
Further, EDSL techniques have spread beyond their origin in 
 the programming languages community.
% 
For example, both Stanford's Parallel Programming Laboratory (PPL) and
the Berkeley Parlab are creating EDSLs as their flagship parallel
programming solutions for domains such as machine learning and
rendering \cite{stanford-ppl}.
  Moreover, EDSLs need not be hosted by esoteric research languages---Intel's
 ArBB embeds an array language in C++ and Berkeley's
Copperhead \cite{copperhead} generates CUDA from simple Python code.

### 2013-10-01 src2/arbb/paperThesis.tex
apa

\newenvironment{mycode}
{% This is the begin code
  \begin{verbatim}}%
  %\footnotesize
  %\noindent
  %\begin{code}\noindent\vspace{-1mm}}
%  \begin{code}\noindent}
{% This is the end code
  \end{verbatim}
  %\end{code} 
}

\newenvironment{inlinecode}
{
  \begin{center}
  \begin{minipage}{4in}
  \begin{mycode}
} 
{
  \end{mycode}
  \end{minipage}
  \end{center}
  \vspace{-1.5ex}
}

@@

% We present our ongoing work on a
In pursuit of this vision, we offer a
% 
 new EDSL implementation, called \accarbb{}, that combines
existing systems, Accelerate \cite{Accelerate} and ArBB \cite{ArBB}, to produce a unified
high-level programming environment equally suited to multicore,
vectorized CPUs, as to GPUs and other accelerators (such as Intel MIC chips \cite{larrabee}).
%
\if{0}
ArBB is based on the RapidMind \cite{RapidMind} system which was an embedded 
DSL targeting multicore CPUs as well as GPUs. ArBB does not \new{yet} include a GPU 
backend like RapidMind, targeting multicore CPUs and their vector 
units. 
\fi{}
\systemname{} is a single EDSL implementation 
with independently optimized backends \new{by different teams; 
namely, the ArBB backend for CPU/MIC, and a CUDA backend for NVIDIA GPU.}
This makes
\systemname{} an appealing platform for fair CPU/GPU comparisons, 
as well as a compelling programming model for single-source portable performance across a range of
parallel architectures, \new{present and future}.
%%To our knowledge, \systemname{} is the first EDSL implementation with
%%independently optimized backends for CPUs and GPUs.  This makes
%%\systemname{} an appealing platform for fair CPU/GPU comparisons, 
%%as well as a compelling programming
%%model for single-source portable performance across a range of
%%parallel architectures.

@@

\vspace{1mm}
\begin{code}
  float f(float x) \{ return (2*x - 1); \}
\end{code}
\noindent
But simply by changing the types, \cde{f} might be lifted to operate on {\em
  expressions} (which, when evaluated, will yield \cde{floats}):

\vspace{1mm}
\begin{code}
  \kw{exp<float>} f(\kw{exp<float>} x) \{
    return (2*x - 1);
  \}
\end{code}

@@

One reason that EDSLs are good for productivity is that the programmer
gains the software engineering benefits of the host language
(object-orientation, higher-order-functions, modules, etc), while not
paying the cost at runtime for additional layers of abstraction or
indirection.  Indeed, the embedded languages for 
performance-oriented EDSLs are often simple, first-order
languages without pointers \cite{wavescript, Accelerate}.

@@

%% dotProd :: Vector Float -> 
%%            Vector Float -> 
%%            Acc (Scalar Float) 
\begin{verbatim}
dotProd (xs :: Vector Float) 
        (ys :: Vector Float) = 
  \kw{let} xs\_ = use xs 
      ys\_ = use ys 
  \kw{in}  fold (+) 0 (zipWith (*) xs\_ ys\_) 
\end{verbatim}

@@

%\vspace{2mm}
\begin{code}
\kw{void} dot\_product(\kw{const} dense<f32>& a, 
                 \kw{const} dense<f32>& b,
                 f32& c) 
\{ 
  c = add\_reduce(a * b);
\}
\end{code}

@@

\vspace{2mm}
\begin{code}
\kw{int} main() 
\{ 
  double a[SIZE];
  double b[SIZE];

  for ( \kw{int} i = 0; i < SIZE; ++i) \{ 
    a[i] = ...; b[i] = ...;
  \}
  dense<f32> va, vb;
  f32 vc;
  bind(va, a, SIZE); 
  bind(vb, b, SIZE); 
  call(dot\_product)(va, vb, &vc); 
  ...
\}
\end{code}

@@

Black-Scholes option pricing is a finance-related benchmark that has been used in similar
DSLs targeting GPUs \cite{Accelerate, nikola}. Since we are re-using the 
Accelerate front-end, we can directly use the Black-Scholes benchmark that 
is shipped with that system.  Figure \ref{fig:blackscholes} shows the
complete code listing for an Accelerate Black-Scholes function which
can be executed on GPUs or any processor targeted by ArBB.

@@

Microsoft Accelerator~\cite{accelerator} is an embedded language with similar 
aspirations as ArBB, that is, to target a diverse range of architectures using the 
same source code. Accelerate can be used from the C\# language or the 
functional F\# language and targets GPUs or of CPUs and their
vector units.
% However, being part of the .Net framework implies limited accessibility 
%for platforms other than Microsoft Windows.

@@

This is a compiler backend research challenge.
The {\em skeletons} method mentioned in
Section~\ref{sec:accelerate-arbb} is one approach to this problem, as
are the optimizations studied in the Copperhead \cite{copperhead} and
Obsidian \cite{obsidian} projects.
%
On the other hand, systems that rely on advanced optimizations
typically suffer to some extent from performance-predictability
problems.  Thus achieving portable, predictable performance on a wide range
of architectures---even for the simplest target languages---will
be the subject of much future work.

>>>>

%% \newenvironment{mycode}
%% {% This is the begin code
%%   \begin{Verbatim}[commandchars=\\\{\}]}%
%%   %\footnotesize
%%   %\noindent
%%   %\begin{code}\noindent\vspace{-1mm}}
%% %  \begin{code}\noindent}
%% {% This is the end code
%%   \end{Verbatim}
%%   %\end{code} 
%% }

%% \newenvironment{inlinecode}
%% {
%%   \begin{center}
%%   \begin{minipage}{4in}
%%   \begin{mycode}
%% } 
%% {
%%   \end{mycode}
%%   \end{minipage}
%%   \end{center}
%%   \vspace{-1.5ex}
%% }

@@

% We present our ongoing work on a
In pursuit of this vision, we offer a
% 
 new EDSL implementation, called \accarbb{}, that combines
existing systems, Accelerate \cite{ACCELERATEDAMP11} and ArBB \cite{ArBB}, to produce a unified
high-level programming environment equally suited to multicore,
vectorized CPUs, as to GPUs and other accelerators (such as Intel MIC chips \cite{larrabee}).
%
\if{0}
ArBB is based on the RapidMind \cite{RapidMind} system which was an embedded 
DSL targeting multicore CPUs as well as GPUs. ArBB does not \new{yet} include a GPU 
backend like RapidMind, targeting multicore CPUs and their vector 
units. 
\fi{}
\systemname{} is a single EDSL implementation 
with independently optimized backends \new{by different teams; 
namely, the ArBB backend for CPU/MIC, and a CUDA backend for NVIDIA GPU.}
This makes
\systemname{} an appealing platform for fair CPU/GPU comparisons, 
as well as a compelling programming model for single-source portable performance across a range of
parallel architectures, \new{present and future}.
%%To our knowledge, \systemname{} is the first EDSL implementation with
%%independently optimized backends for CPUs and GPUs.  This makes
%%\systemname{} an appealing platform for fair CPU/GPU comparisons, 
%%as well as a compelling programming
%%model for single-source portable performance across a range of
%%parallel architectures.

@@

\vspace{1mm}
\begin{Verbatim}[commandchars=\\\{\}]
  float f(float x) \{ return (2*x - 1); \}
\end{Verbatim}
\noindent
But simply by changing the types, \cde{f} might be lifted to operate on {\em
  expressions} (which, when evaluated, will yield \cde{floats}):

\vspace{1mm}
\begin{Verbatim}[commandchars=\\\{\}]
  \kw{exp<float>} f(\kw{exp<float>} x) \{
    return (2*x - 1);
  \}
\end{Verbatim}

@@

One reason that EDSLs are good for productivity is that the programmer
gains the software engineering benefits of the host language
(object-orientation, higher-order-functions, modules, etc), while not
paying the cost at runtime for additional layers of abstraction or
indirection.  Indeed, the embedded languages for 
performance-oriented EDSLs are often simple, first-order
languages without pointers \cite{wavescript, ACCELERATEDAMP11}.

@@

%% dotProd :: Vector Float -> 
%%            Vector Float -> 
%%            Acc (Scalar Float) 
\begin{Verbatim}[commandchars=\\\{\}]
dotProd (xs :: Vector Float) 
        (ys :: Vector Float) = 
  \kw{let} xs\_ = use xs 
      ys\_ = use ys 
  \kw{in}  fold (+) 0 (zipWith (*) xs\_ ys\_) 
\end{Verbatim}

@@

%\vspace{2mm}
\begin{Verbatim}[commandchars=\\\{\}]
\kw{void} dot\_product(\kw{const} dense<f32>& a, 
                 \kw{const} dense<f32>& b,
                 f32& c) 
{ 
  c = add\_reduce(a * b);
}
\end{Verbatim}

@@

\vspace{2mm}
\begin{Verbatim}[commandchars=\\\{\}]
\kw{int} main() 
{ 
  double a[SIZE];
  double b[SIZE];

  for ( \kw{int} i = 0; i < SIZE; ++i) { 
    a[i] = ...; b[i] = ...;
  }
  dense<f32> va, vb;
  f32 vc;
  bind(va, a, SIZE); 
  bind(vb, b, SIZE); 
  call(dot\_product)(va, vb, &vc); 
  ...
}
\end{Verbatim}

@@

Black-Scholes option pricing is a finance-related benchmark that has been used in similar
DSLs targeting GPUs \cite{ACCELERATEDAMP11, NIKOLA}. Since we are re-using the 
Accelerate front-end, we can directly use the Black-Scholes benchmark that 
is shipped with that system.  Figure \ref{fig:blackscholes} shows the
complete code listing for an Accelerate Black-Scholes function which
can be executed on GPUs or any processor targeted by ArBB.

@@

Microsoft Accelerator~\cite{ACCELERATOR} is an embedded language with similar 
aspirations as ArBB, that is, to target a diverse range of architectures using the 
same source code. Accelerate can be used from the C\# language or the 
functional F\# language and targets GPUs or of CPUs and their
vector units.
% However, being part of the .Net framework implies limited accessibility 
%for platforms other than Microsoft Windows.

@@

This is a compiler backend research challenge.
The {\em skeletons} method mentioned in
Section~\ref{sec:accelerate-arbb} is one approach to this problem, as
are the optimizations studied in the Copperhead \cite{copperhead} and
Obsidian \cite{OBSIDIAN-TECHREPORT} projects.
%
On the other hand, systems that rely on advanced optimizations
typically suffer to some extent from performance-predictability
problems.  Thus achieving portable, predictable performance on a wide range
of architectures---even for the simplest target languages---will
be the subject of much future work.

### 2013-10-01 src2/bb/body.tex
apa

\section{Introduction}
\studname{} is a computer science student working in a project involving programming 
robots in a low-level imperative language. However, \studname{} has a budding 
interest in functional programming using Haskell and has read ``The Haskell 
school of expression'' \cite{HUDAK}. He gets the idea to implement 
a language for robot control embedded in Haskell. \studname{} realises that the 
capabilities of the robot hardware are very similar to those of the 
robots that Hudak evaluates graphically in a grid world. There is however 
a very important difference; the embedded language needs to be compiled into 
some form that is understood by the robot. 
%\studname{} in a computer science student with a budding interest in functional 
%programming using Haskell. \studname{} is developing  
%an embedded domain specific language (EDSL) for robot control. Inspired 
%by the book ``the Haskell school of expression'' \cite{HUDAK} but instead of 
%evaluating his robot EDSL, \studname{} wants to compile the language to a form 
%he can run on actual robot hardware. 
%with 
%different motivations, \studname{} wants compile his EDSL for actual physical robots, 
%he sets out.

@@

\subsection{Example programs}

@@

\subsection{Implementation and data structures}

@@

\subsection{Problem statement} 
\emph{Meanwhile in Professor \docname{}'s Office}\newline %\newline

@@

\begin{figure*}
\begin{tabular}{c|c}
\begin{minipage}[]{0.45\columnwidth}
\begin{small}
\begin{verbatim}
spiralIn :: Int -> Program () 
spiralIn 0 = return () 
spiralIn n = do
  replicateM_ 2 $ do
    replicateM_ n move
    turnLeft
  spiralIn (n-1) 
\end{verbatim}
\end{small}
\end{minipage}
\hspace{0.05\columnwidth} 
\begin{minipage}[]{0.45\columnwidth}
\includegraphics[width=\columnwidth]{./bb/spiral1}
\end{minipage}
&
\begin{minipage}[]{0.45\columnwidth}
\begin{small}
\begin{verbatim} 
followWall :: Program () 
followWall = 
  while (return true) $ 
    cond checkLeft sMove $ 
            do turnLeft
               move

checkLeft :: Program BoolE     
checkLeft = do
  TurnLeft
  s <- Sensor
  TurnRight 
  return s
\end{verbatim}
\end{small}
\end{minipage}
\hspace{0.05\columnwidth}
\begin{minipage}[]{0.45\columnwidth}
\includegraphics[width=\columnwidth]{./bb/wall2}
\end{minipage}

\end{tabular}
\caption{The {\tt spiralIn} and {\tt followWall} programs sent from \studname{}
to \docname{} in an email}
\label{fig:programs}
\end{figure*}

\section{Compilation of the monadic robot EDSL} 
\emph{\docname{} receives \studname{}'s latest email}\newline %\newline

@@

\pagebreak
\begin{small} 
\begin{verbatim}
runCompile :: Program a -> Prg
runCompile prg = snd $ compile s prg 
  where 
    s = newNameSupply

@@

\section{Meeting in \docname{}'s office} %\section{Related Work} 
%\emph{Meeting in \docname{}'s office} \newline

@@

\subsection{Related work}

@@

\subsection{Compositionality}

@@

\subsection{The Monad laws}

@@

\section{Composing reifiable monadic languages} 
\label{sec:composing}

@@

\section{Epilogue}

@@

The example language used throughout the story is small and quite
limited. But the technique presented in this paper does scale up to
larger languages and more complicated language constructs.  It is
currently used in the implementation of Obsidian, an EDSL for general
purpose GPU programming \cite{Obsidian}. Obsidian has language
constructs which are considerably more advanced than the robot
language is this paper. One example is the sequential for-loop which
is higher-order (the type parameter {\tt t} is not relevant to the
current discussion):
\begin{small}
\begin{verbatim}
SeqFor :: EWord32 -> (EWord32 -> Program t ())
       -> Program t ()
\end{verbatim}
\end{small}
There does, however, seem to be restrictions on what kind of language
constructs the reification technique presented in this paper can deal
with. For instance, the functions from the {\tt MonadPlus} class has
resisted our attempts at adding them to an EDSL in the same way as
we've added {\tt Return} and {\tt Bind}. The exact limits of the
expressiveness of our method is currently unknown to us.

>>>>

\subsection{Introduction}
\studname{} is a computer science student working in a project involving programming 
robots in a low-level imperative language. However, \studname{} has a budding 
interest in functional programming using Haskell and has read ``The Haskell 
school of expression'' \cite{HUDAK}. He gets the idea to implement 
a language for robot control embedded in Haskell. \studname{} realises that the 
capabilities of the robot hardware are very similar to those of the 
robots that Hudak evaluates graphically in a grid world. There is however 
a very important difference; the embedded language needs to be compiled into 
some form that is understood by the robot. 
%\studname{} in a computer science student with a budding interest in functional 
%programming using Haskell. \studname{} is developing  
%an embedded domain specific language (EDSL) for robot control. Inspired 
%by the book ``the Haskell school of expression'' \cite{HUDAK} but instead of 
%evaluating his robot EDSL, \studname{} wants to compile the language to a form 
%he can run on actual robot hardware. 
%with 
%different motivations, \studname{} wants compile his EDSL for actual physical robots, 
%he sets out.

@@

\subsubsection{Example programs}

@@

\subsubsection{Implementation and data structures}

@@

\subsubsection{Problem statement} 
\emph{Meanwhile in Professor \docname{}'s Office}\newline %\newline

@@

\begin{figure*}
%\begin{tabular}{c|c}
\begin{minipage}[]{0.50\linewidth}
\begin{small}
\begin{verbatim}
spiralIn :: Int -> Program () 
spiralIn 0 = return () 
spiralIn n = do
  replicateM_ 2 $ do
    replicateM_ n move
    turnLeft
  spiralIn (n-1) 
\end{verbatim}
\end{small}
\end{minipage}
\hspace{0.05\linewidth} 
\begin{minipage}[]{0.45\linewidth}
\includegraphics[width=\linewidth]{./bb/spiral1}
\end{minipage}
%&
\begin{minipage}[]{0.50\linewidth}
\begin{small}
\begin{verbatim} 
followWall :: Program () 
followWall = 
  while (return true) $ 
    cond checkLeft sMove $ 
            do turnLeft
               move

checkLeft :: Program BoolE     
checkLeft = do
  TurnLeft
  s <- Sensor
  TurnRight 
  return s
\end{verbatim}
\end{small}
\end{minipage}
\hspace{0.05\linewidth}
\begin{minipage}[]{0.45\linewidth}
\includegraphics[width=\linewidth]{./bb/wall2}
\end{minipage}

%\end{tabular}
\caption{The {\tt spiralIn} and {\tt followWall} programs sent from \studname{}
to \docname{} in an email}
\label{fig:programs}
\end{figure*}

\subsection{Compilation of the monadic robot EDSL} 
\emph{\docname{} receives \studname{}'s latest email}\newline %\newline

@@

%\pagebreak
\begin{small} 
\begin{verbatim}
runCompile :: Program a -> Prg
runCompile prg = snd $ compile s prg 
  where 
    s = newNameSupply

@@

\subsection{Meeting in \docname{}'s office} %\section{Related Work} 
%\emph{Meeting in \docname{}'s office} \newline

@@

\subsubsection{Related work}

@@

\subsubsection{Compositionality}

@@

\subsubsection{The Monad laws}

@@

\subsection{Composing reifiable monadic languages} 
\label{sec:composing}

@@

\subsection{Epilogue}

@@

The example language used throughout the story is small and quite
limited. But the technique presented in this paper does scale up to
larger languages and more complicated language constructs.  It is
currently used in the implementation of Obsidian, an EDSL for general
purpose GPU programming \cite{Obsidian-Expressive}. Obsidian has language
constructs which are considerably more advanced than the robot
language is this paper. One example is the sequential for-loop which
is higher-order (the type parameter {\tt t} is not relevant to the
current discussion):
\begin{small}
\begin{verbatim}
SeqFor :: EWord32 -> (EWord32 -> Program t ())
       -> Program t ()
\end{verbatim}
\end{small}
There does, however, seem to be restrictions on what kind of language
constructs the reification technique presented in this paper can deal
with. For instance, the functions from the {\tt MonadPlus} class has
resisted our attempts at adding them to an EDSL in the same way as
we've added {\tt Return} and {\tt Bind}. The exact limits of the
expressiveness of our method is currently unknown to us.

### 2013-10-01 src2/csort/GPUs.tex
apa

\section{GPUs and CUDA} 
%\FloatBarrier
% This section provides an overview of GPU programming and architecture.

@@

GPUs supporting CUDA are based on a scalable design. The GPU consists of a 
number of so-called multiprocessors (MPs) each consisting of a number of 
processing elements (PEs); sometimes these are referred to as cores or as 
stream processors. An MP also contains a local shared memory through which 
threads running on the PEs can communicate. The key that 
makes the architecture scalable is that a GPU consists of one or more 
of these MPs. This also influences the programming model significantly; since
threads can only communicate in a synchronised manner using the local memory 
(or in a very limited way using atomic operations) the programmer must 
partition the computation in such a way that its communication patterns 
follow this architectural constraint. The GPU also has access to a larger 
memory called the {\em global} or {\em device} memory. The sizes of these 
memories and the number of MPs and PEs per MP vary slightly between 
generations of GPUs. For example, the GPU used in the benchmarks 
(section \ref{sec:Benchmarks}) has a total of 1344 single-precision floating
point CUDA cores and has 2 GB global memory. The very latest GPU architectures 
from NVIDIA slightly digress from what is outlined here but mostly still 
adhere to the same programming model. For exact details refer to \cite{KEPLER}.

>>>>

\subsection{GPUs and CUDA} 
%\FloatBarrier
% This section provides an overview of GPU programming and architecture.

@@

GPUs supporting CUDA are based on a scalable design. The GPU consists of a 
number of so-called multiprocessors (MPs) each consisting of a number of 
processing elements (PEs); sometimes these are referred to as cores or as 
stream processors. An MP also contains a local shared memory through which 
threads running on the PEs can communicate. The key that 
makes the architecture scalable is that a GPU consists of one or more 
of these MPs. This also influences the programming model significantly; since
threads can only communicate in a synchronised manner using the local memory 
(or in a very limited way using atomic operations) the programmer must 
partition the computation in such a way that its communication patterns 
follow this architectural constraint. The GPU also has access to a larger 
memory called the {\em global} or {\em device} memory. The sizes of these 
memories and the number of MPs and PEs per MP vary slightly between 
generations of GPUs. For example, the GPU used in the benchmarks 
(section \ref{sec:CSORTBenchmarks}) has a total of 1344 single-precision floating
point CUDA cores and has 2 GB global memory. The very latest GPU architectures 
from NVIDIA slightly digress from what is outlined here but mostly still 
adhere to the same programming model. For exact details refer to \cite{KEPLER}.

### 2013-10-01 src2/csort/intro.tex
apa

The contributions of this paper are:
\begin{itemize}
\item We provide measurements (section \ref{sec:Benchmarks}) showing
  that counting sort is a competitive algorithm for sorting keys on the GPU,
  outperforming the sorting implementation in the library
  Thrust\cite{THRUST} in many cases.
\item Occurrence sort is shown to be particularly
  suitable for implementing on the GPU (section \ref{sec:occur}) and
  performs well (section \ref{sec:Benchmarks}).
\item The Obsidian implementation of the two sorting algorithms is
  detailed along with the generated CUDA (sections
  \ref{sec:Obsidian} and \ref{sec:parallel}).
\end{itemize}

\subsection{Related work}

>>>>

The contributions of this paper are:
\begin{itemize}
\item We provide measurements (section \ref{sec:CSORTBenchmarks}) showing
  that counting sort is a competitive algorithm for sorting keys on the GPU,
  outperforming the sorting implementation in the library
  Thrust\cite{THRUST} in many cases.
\item Occurrence sort is shown to be particularly
  suitable for implementing on the GPU (section \ref{sec:occur}) and
  performs well (section \ref{sec:CSORTBenchmarks}).
\item The Obsidian implementation of the two sorting algorithms is
  detailed along with the generated CUDA (sections
  \ref{sec:Obsidian} and \ref{sec:parallel}).
\end{itemize}

\subsubsection{Related work}

### 2013-10-01 src2/embarbb/introduction.tex
apa

ArBB is an embedded language, implemented
as a library in the host language C++. The embedding makes 
heavy use of the C++ template machinery. There is also a lower level 
interface to ArBB, the ArBB Virtual Machine C API~\cite{arbbvm}. The low level C API is not
intended for use directly by applications programmers, but rather as a base 
for implementations of alternative ArBB frontends (embeddings)~\cite{ARBB2011}.
The ArBB VM implements the compilation and parallelisation.
Reference~\cite{joelryan} presents Haskell bindings to the ArBB C API as well as initial 
steps towards using ArBB as a backend to the embedded language Accelerate~\cite{accelerate}.

>>>>

ArBB is an embedded language, implemented
as a library in the host language C++. The embedding makes 
heavy use of the C++ template machinery. There is also a lower level 
interface to ArBB, the ArBB Virtual Machine C API~\cite{arbbvm}. The low level C API is not
intended for use directly by applications programmers, but rather as a base 
for implementations of alternative ArBB frontends (embeddings)~\cite{ARBB2011}.
The ArBB VM implements the compilation and parallelisation.
Reference~\cite{joelryan} presents Haskell bindings to the ArBB C API as well as initial 
steps towards using ArBB as a backend to the embedded language Accelerate~\cite{ACCELERATEDAMP11}.

### 2013-10-01 src2/embarbb/programming.tex
apa

The {\tt addconst} function needs to be {\em captured} before it can be executed. 
The term {\em capture} is borrowed from ArBB nomenclature and corresponds to compiling 
the embedded language function into an ArBB function. In EmbArBB, a function is 
captured using the function {\tt capture}; in the C++ embedding, a function is 
captured when it is called for the first time. The same could have been done in 
the Haskell embedding of course, but we chose to make
capture explicit for simplicity.
The {\tt capture} function takes a Haskell function as input and produces an 
opaque typed function identifier as output. The identifier points out the function 
in an environment that is managed by a monad called {\tt ArBB}. To make this 
concrete, Figure~\ref{fig:code1} shows the complete procedure from capturing 
to execution.

@@

%main = 
% withArBB $ 
%  do 
%     f <- capture addconst
%     let x = fromList [0..10 :: Float]      
%     r1 <- liftIO$ new1D 10   
%     execute f (1 :- x)  r1   
%     r <- liftIO$ freeze r1             
%     liftIO$ putStrLn$ show r   
\begin{figure}
\begin{Verbatim}[numbers=left,frame=single] 
main = 
  withArBB $ 
  do 
     f <- capture addconst             
     x <- copyIn $ mkDVector 
                     (V.fromList [1..10 :: Float]) 
                     (Z:.10)
     r1 <- new (Z:.10) 0 
     c <- mkScalar 1      
     execute f (c :- x)  r1              
     r <- copyOut r1              
     liftIO$ putStrLn$ show r
\end{Verbatim}
\caption{The code shows how to capture a function, upload data to ArBB and 
         how to execute the captured function.}
\label{fig:code1}
\end{figure}  
%$
%** Note, I think it would be best to put the code above into a figure and then just refer to it in the discussion, rather than have three versions of it of
%increasing size.

@@

%Matrix multiplication is studied in the benchmarks in section~\ref{sec:Benchmarks}.
The following EmbArBB implementation of matrix-matrix multiplication is used as a benchmark in section~\ref{sec:Benchmarks}:

>>>>

The {\tt addconst} function needs to be {\em captured} before it can be executed. 
The term {\em capture} is borrowed from ArBB nomenclature and corresponds to compiling 
the embedded language function into an ArBB function. In EmbArBB, a function is 
captured using the function {\tt capture}; in the C++ embedding, a function is 
captured when it is called for the first time. The same could have been done in 
the Haskell embedding of course, but we chose to make
capture explicit for simplicity.
The {\tt capture} function takes a Haskell function as input and produces an 
opaque typed function identifier as output. The identifier points out the function 
in an environment that is managed by a monad called {\tt ArBB}. To make this 
concrete, Figure~\ref{fig:EmbArBBcode1} shows the complete procedure from capturing 
to execution.

@@

%main = 
% withArBB $ 
%  do 
%     f <- capture addconst
%     let x = fromList [0..10 :: Float]      
%     r1 <- liftIO$ new1D 10   
%     execute f (1 :- x)  r1   
%     r <- liftIO$ freeze r1             
%     liftIO$ putStrLn$ show r   
\begin{figure}
\begin{Verbatim}[numbers=left,frame=single] 
main = 
  withArBB $ 
  do 
     f <- capture addconst             
     x <- copyIn $ mkDVector 
                     (V.fromList [1..10 :: Float]) 
                     (Z:.10)
     r1 <- new (Z:.10) 0 
     c <- mkScalar 1      
     execute f (c :- x)  r1              
     r <- copyOut r1              
     liftIO$ putStrLn$ show r
\end{Verbatim}
\caption{The code shows how to capture a function, upload data to ArBB and 
         how to execute the captured function.}
\label{fig:EmbArBBcode1}
\end{figure}  
%$
%** Note, I think it would be best to put the code above into a figure and then just refer to it in the discussion, rather than have three versions of it of
%increasing size.

@@

%Matrix multiplication is studied in the benchmarks in section~\ref{sec:Benchmarks}.
The following EmbArBB implementation of matrix-matrix multiplication is used as a benchmark in section~\ref{sec:EmbArBBBenchmarks}:

### 2013-10-01 src2/embarbb/related.tex
apa

\subsection{Related Work} 
\subsubsection{Data.Array.Accelerate} 
Accelerate~\cite{accelerate} is an embedded language for general purpose, 
flat data parallel computations on GPUs.

@@

Repa is compared to EmbArBB in the benchmarks in section~\ref{sec:Benchmarks}.

>>>>

\subsection{Related Work} 
\subsubsection{Data.Array.Accelerate} 
Accelerate~\cite{ACCELERATEDAMP11} is an embedded language for general purpose, 
flat data parallel computations on GPUs.

@@

Repa is compared to EmbArBB in the benchmarks in section~\ref{sec:EmbArBBBenchmarks}.

### 2013-10-01 src2/expressive/applications.tex
apa

The bitonic merger is typically presented as a recursive
construction and we have earlier explored ways to describe
and analyse it in both (our) Ruby and in Lava~\cite{sortsRuby,LavaSorter}.
Here, we consider iterative descriptions using similar combinators.

@@

Table~\ref{fig:table} shows the performance figures for the large sorter
for $5$ different $512-$input small sorter kernels.
{\tt tsort1} is defined above, and is a variant of bitonic sort
that does not require {\tt if} statements to determine the direction
of sorting of pairs, as all comparison operations place the minimum at the
same index as the lower input.
{\tt tsort2} is the same construction, but built with the combinators {\tt ilv2}
and {\tt vee2} that result in the use of one thread per two indices.
{\tt vsort}, the fastest kernel, uses a generalisation of those
two combinators, and again uses one thread per two indices.
{\tt vsort1} is the same construction as {\tt vsort}, but with {\tt ilvVee2} replaced
by {\tt ilvVee1}, and so using one thread per index.
As a reference small kernel implementation, we include a simple hand-coded bitonic sort
that uses one thread per index (see code in Appendix).
None of the kernels has been subject to optimisations related to warp size.

\input{./expressive/table1.tex}

We also recorded the GPU time spent in the calls to the small sorters alone (using the NVIDIA CUDA Visual profiler), see Table~\ref{tab:table2}.
The {\tt tsort1} and {\tt vsort1} kernels, which are built using only pull arrays, 
(and which have one index per thread in the generated code) are
noticeably slower than those that have two indices per thread.
The generation of the latter kernels is made possible by the introduction of 
push arrays.

\input{./expressive/table2.tex}

>>>>

The bitonic merger is typically presented as a recursive
construction and we have earlier explored ways to describe
and analyse it in both (our) Ruby and in Lava~\cite{sortsRuby,LAVASORTER}.
Here, we consider iterative descriptions using similar combinators.

@@

Table~\ref{tab:expressive-table1} shows the performance figures for the large sorter
for $5$ different $512-$input small sorter kernels.
{\tt tsort1} is defined above, and is a variant of bitonic sort
that does not require {\tt if} statements to determine the direction
of sorting of pairs, as all comparison operations place the minimum at the
same index as the lower input.
{\tt tsort2} is the same construction, but built with the combinators {\tt ilv2}
and {\tt vee2} that result in the use of one thread per two indices.
{\tt vsort}, the fastest kernel, uses a generalisation of those
two combinators, and again uses one thread per two indices.
{\tt vsort1} is the same construction as {\tt vsort}, but with {\tt ilvVee2} replaced
by {\tt ilvVee1}, and so using one thread per index.
As a reference small kernel implementation, we include a simple hand-coded bitonic sort
that uses one thread per index (see code in Appendix).
None of the kernels has been subject to optimisations related to warp size.

%\input{./expressive/table1.tex}
\begin{table}
\begin{center}
  \begin{tabular}{| l | r | r | r | r | r | r | }
    \hline
        k                             &  $20$ &  $21$ & $22$& $23$ & $24$ & $24$(CPU)  \\ \hline
    bitonic(CUDA)(512)                &  5 & 12 & 25 & 54 & 117 & 2741   \\      
    tsort1(512)                       &  4 & 10 & 22 & 47 & 102 & 2742    \\
    tsort2(256)                       &  4 & 9 & 23 & 45 & 98 & 2741     \\ 
    vsort1(512)                       &  4 & 10 & 21 & 47 & 102 & 2747    \\ 
    vsort(256)                        &  4 & 9 & 20 & 44 & 96 & 2783    \\ 
    \hline
  \end{tabular}
\end{center}
\caption{Sorting time (ms) for $2^k$ inputs, $20 \leq k \leq 24$.  The GPU used is an NVIDIA GTX480.
Each line shows the result using a particular small $512$ input sort kernel with the
indicated number of threads as {\tt sortSmall} in the CUDA code above. The small merge kernel used
is our generated {\tt bmerge2}, with two indices per thread and $512$ inputs. Memory transfer time (GPU-CPU)
is not shown. In the $2^{24}$ input case using {\tt vsort}, the total sorting time, including
memory transfers was 141 ms.
The rightmost column shows the time taken for the C quicksort function {\tt qsort}, compiled with
{\tt gcc -O3}, to sort the same inputs as those sorted on the GPU on an i7-920 2.8GHz CPU.
} 
\label{tab:expressive-table1}
\end{table}

We also recorded the GPU time spent in the calls to the small sorters alone (using the NVIDIA CUDA Visual profiler), see Table~\ref{tab:expressive-table2}.
The {\tt tsort1} and {\tt vsort1} kernels, which are built using only pull arrays, 
(and which have one index per thread in the generated code) are
noticeably slower than those that have two indices per thread.
The generation of the latter kernels is made possible by the introduction of 
push arrays.

%\input{./expressive/table2.tex}
\begin{table}
\begin{center}
  \begin{tabular}{| l | r | r |   }

    \hline
                              & time & percent \\ \hline
    bitonic(CUDA)             & 30203 & 32.07\\
    tsort1                    & 15823 & 19.72  \\
    vsort1                       & 14955 & 18.76 \\ 
    tsort2                      & 11562 & 15.37    \\ 
    vsort                       & 9228 &  12.67 \\ 
    bitonic (NVIDIA SDK)        & 23815 & 6.55 \\
    \hline
  \end{tabular}
\end{center}
\caption{time: GPU time (in $\mu$s) spent in calls to small sorter kernels 
in the initial phase of the sorter for $2^{24}$ inputs. percent: percentage 
of total GPU time spent in the small sort kernel.
The final line shows the numbers for the bitonic code (for large sorts) that 
is distributed with the the NVIDIA SDK. Its structure also starts with a phase 
of small sorting kernels. It takes 274 ms to sort $2^{24}$ key-value pairs on 
an NVIDIA GTX480 GPU. It could be sped up by using our {\tt vsort} kernel, 
and also by (hand) fusing single ``column'' kernels as we did with {\tt iSwap}, 
although the need to calculate directions of comparators in the bitonic network 
would complicate this. Our CUDA code is simpler because all comparators
point in the same direction in the sorter construction.
} 
\label{tab:expressive-table2}
\end{table}
